Canonical rings of $\bq$-divisors on Minimal Rational Surfaces

### 2015-08-23 surfaces-write-up/arXiv-versions/surface-write-up-arXiv-v1.tex
surfaces: reflect actual arXiv-v1 version

In Section \ref{sec:proj} we prove the following two results bounding
the degree of generators and relations of
stacky canonical rings on $\bp^m$:
\begin{restatable}{thm}{restateeffective}
\label{thm:proj-effective-intro}
Let $D = \sum_{i = 0}^{n} \alpha_i D_i \in \bq \otimes_\bz \di \bp^m$, with $\alpha_i =
\frac{c_i}{k_i} \in \bq_{> 0}$ and each $D_i \in \di \bp^m$ an integral divisor.

@@

\begin{restatable}{thm}{restateproj}
\label{thm:proj-generators-relations}
Let $D = \sum_{i=0}^n \alpha_i D_i \in \bq \otimes_\bz \di \bp^m$
as in Theorem ~\ref{thm:proj-effective-intro}. Write $\ell_i := \lcm
_{j \ne i}(k_j)$ and $a_i := \deg D_i$. Let $\bp^m \cong \proj
\bk[x_0, \ldots, x_m]$. Now suppose that $f_0, \ldots, f_n$ contain
$m$ independent linear polynomials in $x_0, \ldots, x_m$ (with possibly
zero coefficients $\alpha_j$).

Then $R_D$ is generated in degrees at most 
\[
	\sum_{i=0}^n \ell_i a_i
\]

@@

In Section \ref{sec:hirz}, we shift our attention to Hirzebruch surfaces. Recall that for each $m \geq 0$ we can define the
Hirzebruch surface $F_m := \proj(\sym^\bullet \sco_{\bp^1} \oplus
\sco_{\bp^1}(m)),$ or as a quotient of an open subscheme of
$\ba^4$ by a 
weighted action, as is done in the
beginning of Section \ref{sec:hirz}.
Note that Hirzebruch surfaces are regular, normal,
and locally factorial, being $\bp^1$-bundles over
$\bp^1$. Therefore,, the natural map from Cartier
divisors to Weil divisors is an isomorphism.
In other words, all Cartier divisors arise as the zero locus
of a section of some line bundle. We implicitly use this in the
statement of the following theorem. The analogous fact for
projective space was also implicitly used in Theorems
~\ref{thm:proj-effective-intro} and ~\ref{thm:proj-generators-relations}.

\begin{restatable}{thm}{hirzrestate}
\label{thm:hirz-generators-relations}
Let $D = \sum_{i=1}^n \alpha_i D_i \in \bq
\otimes_\bz \di \hirz_m$ 
where $\alpha_i = \frac{c_i}{k_i}$.
Let $u,v,z,w$ be the coordinates for the Hirzebruch surface
$\hirz_m$, as described at the beginning of Section
~\ref{sec:hirz}. For $1 \leq i \leq n$, let $D_i =
V(f_i)$ where $f_i \in \sco(a_i, b_i)$; further, after reordering
the $D_i$ and possibly adding $D_i$ with corresponding $\alpha_i = 0$, suppose $f_1$ and $f_2$ are
independent linear
polynomials in $u$ and $v$ and $f_3$ and $f_4$ are independent
linear polynomials in $w$ and $x$.
Define
\[\ell_{i,j} = \lcm_{h\ne i,j}(k_h).\] 
Then $R_D$ is generated in degrees at most
\[
	\rho = \sum_{i\in \Te(D)} \gcd(\bida_i, \bidb_i)\ell_i +
	\sum_{\substack{i \in \Tp(D) \\	j \in \Tm(D)}} (\bida_i \bidb_j
	- \bida_j \bidb_i) \ell_{i, j}
\]

@@

\begin{rem}
\label{rem:proj-two-points}
Note that we cannot extend this result to the case when
$D$ is supported at two hypersurfaces with arbitrary rational (non-effective)
coefficients in the same manner that O'Dorney does for the
$\bp^1$ case \cite[Section 4]{dorney:canonical}. As will be shown
in Example ~\ref{eg:hyperplane}, the degrees of generation of the
canonical ring of a general $\bq$-divisor supported on two
hyperplanes cannot be bounded so tightly. The two-point
$\bp^1$ result leverages the fact that $\bp^1$ has precisely two
independent coordinates, so that two distinct integral subschemes
cannot repreesnt equations of only $m$ of the $m+1$ coordinates.
\end{rem}

@@

\begin{figure}
\includegraphics{pics/spin-lower-approximations-pic-pics.pdf}
\caption{This figure shows each non-negative best lower
approximation of $\frac{14}{5}.$ Each ``$\bullet$'' denotes a best
lower approximation and each ``$\circ$'' denotes a lattice point
below $5y=14x$ which is not a best lower approximation.  Note that
the non-negative best lower approximations generate the monoid of
lattice points in the first quadrant satisfying  $5y \le 14x$, with
the operation $(a_1, b_1)(a_2, b_2)\mapsto (a_1 + a_2, b_1 + b_2)$.}
\label{fig:s14/5-lattice}
\end{figure}

>>>>

In Section \ref{sec:proj} we prove the following two results bounding
the degree of generators and relations of
stacky canonical rings on $\bp^m$. The first applies
to effective $\bq$-divisors and the second applies to arbitrary
$\bq$-divisors.
\begin{restatable}{thm}{restateeffective}
\label{thm:proj-effective-intro}
Let $D = \sum_{i = 0}^{n} \alpha_i D_i \in \bq \otimes_\bz \di \bp^m$, with $\alpha_i =
\frac{c_i}{k_i} \in \bq_{> 0}$ and each $D_i \in \di \bp^m$ an integral divisor.

@@

\begin{restatable}{thm}{restateproj}
\label{thm:proj-generators-relations}
Let $D = \sum_{i = 0}^n \alpha_i D_i \in \bq \otimes_\bz \di \bp^m$
with $\alpha_i = \frac{c_i}{k_i}\in \bq$. Write $\ell_i := \lcm
_{j \ne i}(k_j)$ and $a_i := \deg D_i$. Let $\bp^m \cong \proj
\bk[x_0, \ldots, x_m]$ and let each $D_i = V(f_i)$ with $f_i \in \bk[x_0, \ldots, 
x_m]$. Now suppose that $\{f_0, \ldots, f_n\}$ contains $m + 1$ independent 
linear polynomials in $x_0, \ldots, x_m$ {\rm(}with corresponding 
$\alpha_j$ possibly zero{\rm)}.

Then $R_D$ is generated in degrees at most 
\[
	\sum_{i = 0}^n \ell_i a_i
\]

@@

In Section \ref{sec:hirz}, we shift our attention to Hirzebruch surfaces. Recall that for each $m \geq 0$ we can define the
Hirzebruch surface $F_m := \proj(\sym^\bullet \sco_{\bp^1} \oplus
\sco_{\bp^1}(m)),$ or as a quotient of an open subscheme of
$\ba^4$ by a 
weighted action, as is done in the
beginning of Section \ref{sec:hirz}.
Note that Hirzebruch surfaces are regular, normal,
and locally factorial, being $\bp^1$-bundles over
$\bp^1$. Therefore, the natural map from Cartier
divisors to Weil divisors is an isomorphism.
In other words, all Cartier divisors arise as the zero locus
of a section of some line bundle. We implicitly use this in the
statement of the following theorem. The analogous fact for
projective space was also implicitly used in Theorems
~\ref{thm:proj-effective-intro} and ~\ref{thm:proj-generators-relations}.

\begin{restatable}{thm}{hirzrestate}
\label{thm:hirz-generators-relations}
Let $D = \sum_{i = 1}^n \alpha_i D_i \in \bq
\otimes_\bz \di \hirz_m$ where $\alpha_i = \frac{c_i}{k_i} \in \bq$.
Write $\ell_{i,j} := \lcm_{h\ne i,j}(k_h)$.
Let $u, v, z, w$ be the coordinates for the Hirzebruch surface
$\hirz_m$, as described at the beginning of Section
~\ref{sec:hirz} and define $\Te, \Tp$, and $\Tm$ as in Definition ~\ref{defn:t-defn}. Let each $D_i = V(f_i)$ where $f_i \in \sco(a_i, b_i)$.
Now suppose that $\{f_1, \ldots, f_n\}$ contains two independent 
linear polynomials in $u, v$ and two independent linear polynomials in
$w, x$ {\rm(}with corresponding $\alpha_j$ possibly zero{\rm)}.

Then $R_D$ is generated in degrees at most
\[
	\rho = \sum_{i\in \Te(D)} \gcd(\bida_i, \bidb_i)\ell_i +
	\sum_{\substack{i \in \Tp(D) \\	j \in \Tm(D)}} (\bida_i \bidb_j
	- \bida_j \bidb_i) \ell_{i, j}
\]

@@

\begin{rem}
\label{rem:proj-two-points}
Note that we cannot extend this result to the case when
$D$ is supported at two hypersurfaces with arbitrary rational (non-effective)
coefficients in the same manner that O'Dorney does for the
$\bp^1$ case \cite[Section 4]{dorney:canonical}. As will be shown
in Example ~\ref{eg:hyperplane}, the degrees of generation of the
canonical ring of a general $\bq$-divisor supported on two
hyperplanes cannot be bounded so tightly. The two-point
$\bp^1$ result leverages the fact that $\bp^1$ has precisely two
independent coordinates, so that two distinct integral subschemes
cannot represent equations of only $m$ of the $m+1$ coordinates.
\end{rem}

@@

\begin{figure}
\includegraphics{spin-lower-approximations-pic-pics.pdf}
\caption{This figure shows each non-negative best lower
approximation of $\frac{14}{5}.$ Each ``$\bullet$'' denotes a best
lower approximation and each ``$\circ$'' denotes a lattice point
below $5y=14x$ which is not a best lower approximation.  Note that
the non-negative best lower approximations generate the monoid of
lattice points in the first quadrant satisfying  $5y \le 14x$, with
the operation $(a_1, b_1)(a_2, b_2)\mapsto (a_1 + a_2, b_1 + b_2)$.}
\label{fig:s14/5-lattice}
\end{figure}
